---
title: A Crash Course on PostgreSQL for R Users
subtitle: University of Arizona - RezBaz 2021
authors: Mauricio "Pachá" Vargas S.
date: 2020-05-18
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["xaringan-themer.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: ""
      ratio: 16:9
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
xaringanthemer::style_mono_dark()
```

# Motivation

* If you have your workflows that are well organized, and still you create a lot of CSV or RDS files to share periodical data then you can benefit from using PostgreSQL.
* PostgreSQL and MariaDB are the database engines that I like, but this still applies to MySQL and even to propietary databases such as Oracle Database or SQL Server.
* Big disclaimer: Oracle Database and SQL Server are going to be much harder to configure!

---

# Create your own Postres server !

* Create a 1GB RAM + 25GB disk VPS by using the [Supabase Postgres Image](https://marketplace.digitalocean.com/apps/supabase-postgres) from Digital Ocean Marketplace. 
* You can even explore the code on [GitHub](https://github.com/supabase/postgres/).
* If you don't have an account on DigitalOcean, use my [referral link](https://m.do.co/c/6119f0430dad) to have free credits.

---

# Nycflights13

* Assume that you work at NYC airport and you are in charge of updating the `flights` table every day.
* Instead of creating one CSV file per day with flight records, the process can be better organized by appending new records to an existing SQL table.

---

Let's start with the `flights` table from the quite popular `nycflights13` package:
```{r,eval=TRUE}
library(nycflights13)
flights
```

---

* `nycflights13` provides different tables besides `flights`
* You can explore the details in the superb book [R4DS](https://r4ds.had.co.nz/relational-data.html).

---

# SQL indexes

* *Skipping a lot of details.*
* SQL creates copies of the tables that contain just a few columns which help to identify observations.
* This is done in order to be able to filter those observations in a faster way that without an index.

---

For example, in the `month` table, the year column is really useful to create an index that boosts the speed of SQL operations similar to this `dplyr` statement:
```{r, eval=FALSE}
library(dplyr)
flights %>% filter(month == 3)
```

---

To create indexes, let's start differentiating between unique and non-unique indexes:
```{r}
unique_index <- list(
    airlines = list("carrier"),
    planes = list("tailnum")
  )

index <- list(
    airports = list("faa"),
    flights = list(
      c("year", "month", "day"), "carrier", "tailnum", "origin", "dest"
    ),
    weather = list(c("year", "month", "day"), "origin")
  )
```

---

# SQL connections

* We need at least this information in order to connect to a Postgres database:
  * DB name: `postgres`
  * DB host: `123.456.789`
  * DB user: `postgres`
  * DB password: `S4cr3tP4ssw0rd007`
  * DB port: `5432`.
  
---

* Databases information such as passwords should be stored in your `~/.Rprofile` and **never** in your scripts! 
* You can type `usethis::edit_r_environ()` and edit your R profile to add the credentials and then restart your R session to apply the changes.

---

To access and start creating tables to better organize your data, you need to open and close connections, and the `RPostgres` package is really useful to do that and more:

```{r, eval = FALSE}
library(RPostgres)

fun_connect <- function() {
  dbConnect(
    Postgres(),
    dbname = Sys.getenv("tutorial_db"),
    user = Sys.getenv("tutorial_user"),
    password = Sys.getenv("tutorial_pass"),
    host = Sys.getenv("tutorial_host")
  )
}
```

---

By default, PostgreSQL uses the `public` schema in a databases, which you'll write to. You can explore the existing tables in the `public` schema of `postgres` database by running:

```{r ,eval = FALSE}
schema <- "public"

conn <- fun_connect()
remote_tables <- dbGetQuery(
  conn,
  sprintf("SELECT table_name 
          FROM information_schema.tables 
          WHERE table_schema='%s'", schema)
)
dbDisconnect(conn)

remote_tables <- as.character(remote_tables$table_name)
remote_tables
```

Because this is an example with a freshly created server just for the tutorial, the result is no existing tables in the database.

---

# Creating tables

Now you should be ready to start writing the tables from `nycflights13` to the database:

```{r, eval = FALSE}
local_tables <- utils::data(package = "nycflights13")$results[, "Item"]
tables <- setdiff(local_tables, remote_tables)

for (table in tables) {
  df <- getExportedValue("nycflights13", table)
  message("Creating table: ", table)
  table_name <- table
  conn <- fun_connect()
  copy_to(
    conn, df, table_name,
    unique_indexes = unique_index[[table]], indexes = index[[table]], temporary = FALSE
  )
  dbDisconnect(conn)
}
```

This approach prevents errors in case that a table already exists. Also, it prevents a bigger error that would be to overwrite an existing table instead of appending new observations.

---

# Accessing tables

You can use `dplyr` in a very similar way to what you would do with a local `data.frame`. Let's say you need to count the observations per month and save the result locally:

```{r, eval = FALSE}
conn <- fun_connect()
obs_per_month <- tbl(conn, "flights") %>% 
  group_by(year, month) %>% 
  count() %>% 
  collect()
dbDisconnect(conn)

obs_per_month
```

---

* The previous chunk uses the `collect()` function, which stores the result of the query as a local `data.frame`, which can be used with `ggplot2`, saved as CSV with `readr`, etc.
* We even fit a regression with this data!

---

# Appending new records

Let's say you have a table `flights14` with records for the year 2014. To append those records to the existing `flights` table in the database, you need to run:

```{r, eval=FALSE}
conn <- fun_connect()
dbWriteTable(conn, "flights", flights14, append = TRUE, overwrite = FALSE, row.names = FALSE)
dbDisconnect(conn)
```

---

# Appending large tables

Let's say your `flights14` contains 12,345,678 rows. A better approach than writing the full table at once is to divide it in parts of 2,500,000 rows (or other smaller number).

---

## Using base R

```{r, eval=FALSE}
N <- 2500000
divisions <- seq(N, N * ((nrow(flights14) %/% N) + 1), by =  N)
    
flights14_2 <- list()
    
for (j in seq_along(divisions)) {
  if (j == 1) {
    flights14_2[[j]] <- flights14[1:divisions[j],]
  } else {
    flights14_2[[j]] <- flights14[(divisions[j-1] + 1):divisions[j]]
  }
}
    
rm(flights14)
    
for (j in seq_along(divisions)) {
  message(sprintf("Writing fragment %s of %s", j, length(divisions)))
  conn <- fun_connect()
  dbWriteTable(conn, "flights", flights14_2[[j]],
    append = TRUE, overwrite = FALSE, row.names = FALSE)
  dbDisconnect(conn)
}
```

---

## Using the Tidyverse

```{r, eval=FALSE}
library(tidyverse)

N <- 2500000

flights14 <- flights14 %>%
  mutate(p = floor(row_number() / N) + 1) %>% 
  group_by(p) %>% 
  nest() %>% 
  ungroup() %>% 
  select(data) %>% 
  pull()

map(
  seq_along(flights14),
  function(x) {
    message(sprintf("Writing fragment %s of %s", x, length(flights14)))
    conn <- fun_connect()
    dbWriteTable(conn, "flights", flights14_2[[x]],
      append = TRUE, overwrite = FALSE, row.names = FALSE)
    dbDisconnect(conn)
  }
)
```

---

# Questions?
